Learning to Synthesize Rules for Network Intrusion Detectors

### 2020-02-20 paper/paper.tex
pass on section 2

Figure~\ref{fig:synflood-example} shows an example rule of
Suricata~\cite{suricata}, a popular open-source \nids{} maintained by
the Open Information Security Foundation (OISF)~\cite{oisf}. A
Suricata rule is divided in three parts---action, header, and rule
options~\cite{suri-rule-format}. The action part appears as the first
word in the rule description. An action denotes the task that needs to
be executed if the rule pattern is satisfied. In this example, a
message will be sent to system administrators if the rule pattern is
satisfied. The header comes after the action in the rule
description. It restricts the information flow covered by the
rule. For this rule, the header is \CodeIn{tcp \$HOME\_NET any ->
  \$EXTERNAL\_NET any}. It instructs Suricata to inspect \CodeIn{tcp}
traffic flowing from any port in the home network to any other address
outside the home network. The variables \CodeIn{\$HOME\_NET} and
\CodeIn{\$EXTERNAL\_NET} are configurable. The rule options come after
the header. It is a semi-colon-separated sequence of key-value
pairs. Rule options serve to document the analyzed traffic (\eg{},
``msg'', ``classtype'', and ``sid'') and to characterize the attack
pattern (\eg, ``flags'', ``threshold'', and ``content''). \tname{}
produces options of the second kind, which are \emph{relevant} to
capturing the attack.

\begin{table}[t!]
  \caption{\label{table:rules}Example options of Suricata.}  
  \centering
  \begin{tabular}{clp{5.5cm}}
    \toprule
    \multicolumn{1}{c}{\#} & \multicolumn{1}{c}{Name} &  \multicolumn{1}{c}{Description}\\
    \midrule     
    1 & dsize & matches on the size of the packet payload\\
    2 & itype &  matches on a specific ICMP type\\
    3 & icode & matches on a specific ICMP code\\
    4 & icmp\_seq  & checks a ICMP sequence number\\
    5 & icmp\_id & matches on specific ICMP id-values\\
    6 & window & checks the size of the TCP window\\
    7 & flags & matches on TCP flags\\
    8 & fragbits & checks if the fragmentation or reserved bits are set in the IP header\\
    9 & threshold & controls the rule’s alert frequency\\
    10 & content & checks if argument is present on the payload\\
    \bottomrule
  \end{tabular}
\end{table}

Table~\ref{table:rules} shows a sample of the options of
Suricata. Column ``\#'' shows the option number, column ``Name'' shows
the name of the option, and column ``Description'' presents a short
description of the option purpose. We found that the option
\CodeIn{content} is particularly prevalent in rules from public
databases\MyComment{ (see Figure~\ref{fig:distribution-contents})}. This
option takes a sequence of bytes as argument (\eg{},
\CodeIn{content:''SELECT ''}) and checks if that sequence is present
in the payload of the message. Detailed description of the Suricata
rules can be found elsewhere~\cite{suri-rule-format}. It is worth
noting that Suricata supports a total of \totoptions\ distinct options
covering \numproto\ distinct protocols. A total of 9 of these options
are for documentation.\MyComment{For instance, the purpose of rule
  option \CodeIn{msg} is to print on the output an informative message
  indicating what kind of intrusion has been observed.}
Figure~\ref{fig:distribution-rules-protocol} shows the distributions
of these options per protocol. The sum of the number of protocol
options is higher than \totoptions\ as some options are shared across
different protocols. \tname{} does not make distinction among protocols.

%\begin{wrapfigure}[14]{r}{0.5\textwidth}
\begin{figure}[t!]
  \centering
%  \vspace{-7ex}
  \scalebox{0.85}{
    \begin{tikzpicture}
      \begin{axis}[
          ybar stacked,
          enlargelimits=0.15,
          legend style={at={(0.5,-0.15)},
            anchor=north,legend columns=-1},
          %          ylabel={\#rules},
          symbolic x coords={dns, ftp, http, icmp, ip, smb, smtp, ssh,
          tcp, tls, udp},
          xtick=data,
          xticklabel style={rotate=45},          
          nodes near coords,
          every node near coord/.append style={font=\scriptsize},        
          nodes near coords align={vertical},
        ]
        \addplot coordinates {(dns,10) (ftp,11) (http,20) (icmp,18)(ip,16) (smb,13) (smtp,11) (ssh, 10) (tcp, 32) (tls,18) (udp, 18)};
%        \legend{relevant, irrelevant}
      \end{axis}
    \end{tikzpicture}
  }
  \vspace{-2ex}
  \caption{\label{fig:distribution-rules-protocol}Number of relevant
    options per protocol supported by Suricata.}
%\end{wrapfigure}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[scale=0.28]{figs/ICMP-packet-structure.png}
\caption{Layout of an ICMP packet.}
\label{fig:icmp-packet-layout}
\end{figure}

Rule options are often associated with specific fields of network
packets. Let's consider the ICMP protocol as an
example. Figure~\ref{fig:icmp-packet-layout} shows the layout of an
ICMP packet. The packet fields ``Flags'' and ``Time to Live'',
declared in the IP header, are mapped to the options \CodeIn{flagbits}
and \CodeIn{ttl}, respectively. Likewise, the fields ``Type'' and
``Code'', declared in the ICMP header, are mapped to the options
\CodeIn{itype} and \CodeIn{icode}. (Table~\ref{table:rules} lists some
example rule options.) The values of the options ``ttl'', ``itype'',
and ``icode'' are the same as the value of the corresponding fields in
the ICMP packet. For the field ``Flags'' the value of the
corresponding option is encoded in the bitvector. For example, if the
``Flags'' field stores the value 1010, the corresponding rule option
will be \CodeIn{flagbits:RM}, as each position of the vector encodes a
different character in the \CodeIn{flagbits} rule option. The value of
several options can be determined through transformation functions, as
those described above. One expection to this is the option
\CodeIn{threshold}, which is typically used in denial-of-service
attacks. That option operates over sequence of messages as opposed to
a single message. Consequently, it is not possible to determined its
values from a single message. More precisely, to trigger an action on
a rule containing that option a \nids\ tool needs to monitor several
messages.

@@

\begin{figure}[h!]
  \lstinputlisting[language=C,numbers=none,keywords={flags,threshold}]{synflood.suricata}
  \caption{Suricata rule for SYN Flood attack.}
  \label{fig:synflood-example}
\end{figure}

>>>>

Figure~\ref{fig:synflood-example} shows an example rule of
Suricata~\cite{suricata}, a popular open-source \nids{} maintained by
the Open Information Security Foundation (OISF)~\cite{oisf}.

\vspace{-1ex}
\begin{figure}[h!]
  \lstinputlisting[language=C,numbers=none,keywords={flags,threshold}]{synflood.suricata}
  \vspace{-2ex}
  \caption{Suricata rule for SYN Flood attack.}
  \label{fig:synflood-example}
\end{figure}
\vspace{-2ex}

A
Suricata rule is divided in three parts---action, header, and rule
options~\cite{suri-rule-format}. The action part appears as the first
word in the rule description. An action denotes the task that needs to
be executed if the rule pattern is satisfied.
In this example, a
message will be sent to system administrators if the rule pattern is
satisfied. The header comes after the action in the rule
description. It restricts the information flow covered by the
rule. For this rule, the header is \CodeIn{tcp \$HOME\_NET any ->
  \$EXTERNAL\_NET any}. It instructs Suricata to inspect \CodeIn{tcp}
traffic flowing from any port in the home network to any other address
outside the home network. The variables \CodeIn{\$HOME\_NET} and
\CodeIn{\$EXTERNAL\_NET} are configurable. The rule options come after
the header. It is a semi-colon-separated sequence of key-value
pairs. Rule options serve to document the analyzed traffic (\eg{},
``msg'', ``classtype'', and ``sid'') and to characterize the attack
pattern (\eg, ``flags'', ``threshold'', and ``content''). \tname{}
produces options of the second kind, which are \emph{relevant} to
capturing the attack.

\vspace{-1ex}
\begin{table}[h!]
  \footnotesize
  \caption{\label{table:rules}Example options of Suricata.}
  \vspace{-2ex}
  \centering
  \begin{tabular}{clp{5.5cm}}
    \toprule
    \multicolumn{1}{c}{\#} & \multicolumn{1}{c}{Name} &  \multicolumn{1}{c}{Description}\\
    \midrule     
    1 & dsize & matches on the size of the packet payload\\
    2 & itype &  matches on a specific ICMP type\\
    3 & icode & matches on a specific ICMP code\\
    4 & icmp\_seq  & checks a ICMP sequence number\\
    5 & icmp\_id & matches on specific ICMP id-values\\
    6 & window & checks the size of the TCP window\\
    7 & flags & matches on TCP flags\\
    8 & fragbits & checks if the fragmentation or reserved bits are set in the IP header\\
    9 & threshold & controls the rule’s alert frequency\\
    10 & content & checks if argument is present on the payload\\
    \bottomrule
  \end{tabular}
\end{table}
%\vspace{-1ex}

Table~\ref{table:rules} shows a sample of the options of
Suricata. Column ``\#'' shows the option number, column ``Name'' shows
the name of the option, and column ``Description'' presents a short
description of the option purpose. We found that the option
\CodeIn{content} is particularly prevalent in rules from public
databases\MyComment{ (see Figure~\ref{fig:distribution-contents})}. This
option takes a sequence of bytes as argument (\eg{},
\CodeIn{content:''SELECT ''}) and checks if that sequence is present
in the payload of the message. Detailed description of the Suricata
rules can be found elsewhere~\cite{suri-rule-format}. It is worth
noting that Suricata supports a total of \totoptions\ distinct options
covering \numproto\ distinct protocols.

%% A total of 9 of these options
%% are for documentation.\MyComment{For instance, the purpose of rule
%%   option \CodeIn{msg} is to print on the output an informative message
%%   indicating what kind of intrusion has been observed.}
%% Figure~\ref{fig:distribution-rules-protocol} shows the distributions
%% of these options per protocol. The sum of the number of protocol
%% options is higher than \totoptions\ as some options are shared across
%% different protocols. \tname{} does not make distinction among protocols.

%% %\begin{wrapfigure}[14]{r}{0.5\textwidth}
%% \begin{figure}[t!]
%%   \centering
%% %  \vspace{-7ex}
%%   \scalebox{0.85}{
%%     \begin{tikzpicture}
%%       \begin{axis}[
%%           ybar stacked,
%%           enlargelimits=0.15,
%%           legend style={at={(0.5,-0.15)},
%%             anchor=north,legend columns=-1},
%%           %          ylabel={\#rules},
%%           symbolic x coords={dns, ftp, http, icmp, ip, smb, smtp, ssh,
%%           tcp, tls, udp},
%%           xtick=data,
%%           xticklabel style={rotate=45},          
%%           nodes near coords,
%%           every node near coord/.append style={font=\scriptsize},        
%%           nodes near coords align={vertical},
%%         ]
%%         \addplot coordinates {(dns,10) (ftp,11) (http,20) (icmp,18)(ip,16) (smb,13) (smtp,11) (ssh, 10) (tcp, 32) (tls,18) (udp, 18)};
%% %        \legend{relevant, irrelevant}
%%       \end{axis}
%%     \end{tikzpicture}
%%   }
%%   \vspace{-2ex}
%%   \caption{\label{fig:distribution-rules-protocol}Number of relevant
%%     options per protocol supported by Suricata.\Mar{not sure if this
%%       is really helpful.}}
%% %\end{wrapfigure}
%% \end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[scale=0.28]{figs/ICMP-packet-structure.png}
\vspace{-2ex}
\caption{Layout of an ICMP packet.}
\label{fig:icmp-packet-layout}
\end{figure}
%\vspace{-2ex}

Rule options are often associated with specific fields of network
packets. Let's consider the ICMP protocol as an
example. Figure~\ref{fig:icmp-packet-layout} shows the layout of an
ICMP packet. The packet fields ``Flags'' and ``Time to Live'',
declared in the IP header, are mapped to the options \CodeIn{flagbits}
and \CodeIn{ttl}, respectively. Likewise, the fields ``Type'' and
``Code'', declared in the ICMP header, are mapped to the options
\CodeIn{itype} and \CodeIn{icode}. The values of the options ``ttl'',
``itype'', and ``icode'' are the same as the value of the
corresponding fields in the ICMP packet. For the field ``Flags'' the
value of the corresponding option is encoded in a bitvector. For
example, if the ``Flags'' field stores the value 1010, the
corresponding rule option will be \CodeIn{flagbits:RM}, as each
position of the vector encodes a different character in the
\CodeIn{flagbits} rule option. The value of several options can be
determined through transformation functions, as those described
above. One exception to this is the option \CodeIn{threshold}, which
is typically used in denial-of-service attacks. That option operates
over sequence of messages as opposed to a single
message. Consequently, it is not possible to determined its values
from a single message.

%% More precisely, to trigger an action on a rule
%% containing that option a \nids\ tool needs to monitor several
%% messages.

@@
